---
title: "Final project"
author: "Group 5"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(vroom)
library(tidyquant)
library(spotifyr)
library(lubridate)
library(scales)
library(patchwork)
library(corrplot)
```


Thousands of new songs are released every year but very few make it to the top records. We decided to embark on a journey to discover what are the different characteristics of hit songs to help artists maximize their chances of launching a successful song.

We started with with all the songs that have been on the Top 200 Weekly (Global) charts of Spotify in 2020 & 2021 (https://www.kaggle.com/datasets/sashankpillai/spotify-top-200-charts-20202021). The dataset contains features such as danceability, liveness, energy, loudness, chord etc.


```{r}
#loading data and fixing column names using janitor
spotify_data <- vroom("data/spotify.csv", show_col_types = FALSE) %>% 
  clean_names()
```


```{r}
#check NA values for the release date
sum(is.na(spotify_data$`Release Date`))
```

```{r}
#Clean data
spotify_data_clean <- spotify_data %>%
  dplyr::select(-index) %>%  #remove index column
  dplyr::select(-artist_followers,-streams, -song_id, everything()) %>%#move some not-so-important variables to the end. 
  mutate(
    release_date = mdy(release_date),#convert release date from char to date format
    release_month = month(release_date, label=TRUE, abbr = TRUE),
    release_day = day(release_date),
    duration_min = duration_ms / 60000
  ) %>%
  filter(!is.na(release_date))#17 values had a different format than month-day-year. We remove these and the previous NA values.
```

Distribution of songs per date

We want to understand what's the best time to release a song. We start by looking at what's the month in which most top songs were released. Between 2020-2021, most top songs were released in March, May and June. 

```{r}
monthly_distrib <- spotify_data_clean %>%
  group_by(release_month) %>%
  summarize(
    count = n()
  ) 

monthly_distrib <- monthly_distrib %>%
  mutate(top3 = rank(-count) %in% 1:3)
  
ggplot(data = monthly_distrib) +
  aes(x = release_month, 
      y = count, 
      fill= top3) +
  geom_bar(stat = 'identity', width=0.8) + 
  labs(
    x = "Month of release",
    y = "Number of songs",
    title = "Most hits were released in March, May and June"
  ) +
  scale_fill_manual(values = c("TRUE" = "#1db954", "FALSE" = "light grey")) +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.position="none")
```
Next we look at the date of the month. We can also notice that most hits were released on the days of 6th, 10th and 13th.

Looking at the two graphs, we can see that 13th of May is the day when most songs were published. Therefore, publishing a song around 13th of May can increase chances of becoming a hit song. However, we must do a more detailed analysis between concluding this.

```{r}
day_distrib <- spotify_data_clean %>%
  group_by(release_day) %>%
  summarize(
    count = n()
  ) 

day_distrib <- day_distrib %>%
  mutate(top3 = rank(-count) %in% 1:3)

ggplot(data = day_distrib) + 
  aes(x = release_day, y = count, fill = top3) +
  geom_bar(stat = 'identity', width=0.8, aes(xmin=1)) + 
  labs(
    x = "Day of release",
    y = "",
    title = "Most hits were released on the 6th, 10th and 13th"
  ) +
  scale_fill_manual(values = c("TRUE" = "#1db954", "FALSE" = "light grey")) +
  coord_cartesian(xlim =c(1, 31)) +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.position="none")

```

How long should a song be?

We want to explore how duration can impact the popularity of a song. Looking at the graph, we can see that popularity increases with duration and most songs are between 2.5 and 3 minutes long. This is not surprising. Very short songs fail to establish a connection between the artist and listener.

```{r}
#Relation between song popularity and duration

ggplot(data = spotify_data_clean) +
  aes(x = duration_min, y = popularity) +
  geom_point() +
  geom_smooth(color="#1db954", se=FALSE) +
  theme_bw() +
  labs(
    x = "Duration (in min)",
    y = "Popularity",
    title = "Most hits are between 2.5 to 3 minutes long"
  )

```

There is no significant correlation between highest song position in the charts/how long it stayed in the top200 chart and its characteristics like danceability, energy and so on, so it's unlikely that any of those are predictors of whether the song will be a hit, however, we can explore on how those metrics interact among themselves. 

```{r}
res <- cor(spotify_data_clean %>% select(highest_charting_position, number_of_times_charted, popularity, danceability, energy, loudness, speechiness, acousticness, liveness, tempo, duration_ms,valence, streams, artist_followers))
round(res, 2)

corrplot(res, method = "ellipse", type = "upper", color = "green", tl.col = 'black',  tl.srt = 45, col=colorRampPalette(c("#287DC7","white","#1DB954"))(200))
```

The resulting graphs show that loud songs tend to be also the most energetic ones. Acoustic ones, on the other hand, tend to be quieter, less energetic and harder to dance to. Finally, there is positive correlation between how emotional is the song and its danceability, loudness  and energy - so the most intense songs might be clubs' biggest hits.

```{r}
#finding the top 20 genre
data1 <- spotify %>% 
  group_by(Genre) %>% 
  summarise(occurances = n()) %>% 
  slice_max(order_by = occurances, n=20)

ggplot(data1, aes(y=fct_reorder(Genre,(occurances)), x=occurances)) + geom_col(fill='#2EE254', width=0.8)+
  theme(legend.position = "none") +
  labs(title = "Top 20 Genre based on Popularity",
       y = "Genre", 
       x="Popularity")+
  theme_bw()  
```

The graph shows that dance pop is the most popular genre among the songs

```{r}
#finding the top 20 artist
data1 <- spotify %>% 
  group_by(Artist) %>% 
  summarise(occurances = n()) %>% 
  slice_max(order_by = occurances, n=20)

ggplot(data1, aes(y=fct_reorder(Artist,(occurances)), x=occurances)) + geom_col(fill='#2EE254', width=0.8)+
  theme(legend.position = "none") +
  labs(title = "Top 20 Artist based on Popularity",
       y = "Artist", 
       x="Popularity")+
  theme_bw()  

```

Taylor Swift is the most popular artist 

```{r}
#finding the relation between valence and popularity of the track
ggplot(spotify, aes(y=Valence,x=Popularity)) + geom_col(fill='#2EE254', width=0.8)+
  theme(legend.position = "none") +
  labs(title = "Most popular songs happen to be sad songs",
       subtitle="Valence indicates the mood of the song",
       y = "Happiness of the song", 
       x="Popularity")+
  theme_bw() 

```

From the above visualization, we can see that the most popular songs have very low valency. Low valency states that the songs have a sad vibe associated with it. So popular songs have a sad vibe associated to it and that explains why taylor swift is the most popular artist.


