---
title: "Final project"
author: "Your name goes here"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(vroom)
library(tidyquant)
library(spotifyr)
library(lubridate)
library(scales)
library(patchwork)
```


Thousands of new songs are released every year but very few make it to the top records. We decided to embark on a journey to discover what are the different characteristics of hit songs to help artists maximize their chances of launching a successful song.

We started with with all the songs that have been on the Top 200 Weekly (Global) charts of Spotify in 2020 & 2021 (https://www.kaggle.com/datasets/sashankpillai/spotify-top-200-charts-20202021). The dataset contains features such as danceability, liveness, energy, loudness, chord etc.


```{r}
#loading data and fixing column names using janitor
spotify_data <- vroom("data/spotify.csv", show_col_types = FALSE) %>% 
  clean_names()
```


```{r}
#check NA values for the release date
sum(is.na(spotify_data$`Release Date`))
```

```{r}
#Clean data
spotify_data_clean <- spotify_data %>%
  dplyr::select(-index) %>%  #remove index column
  dplyr::select(-artist_followers,-streams, -song_id, everything()) %>%#move some not-so-important variables to the end. 
  mutate(
    release_date = mdy(release_date),#convert release date from char to date format
    release_month = month(release_date, label=TRUE, abbr = TRUE),
    release_day = day(release_date),
    duration_min = duration_ms / 60000
  ) %>%
  filter(!is.na(release_date))#17 values had a different format than month-day-year. We remove these and the previous NA values.
```

Distribution of songs per date

We want to understand what's the best time to release a song. We start by looking at what's the month in which most top songs were released. Between 2020-2021, most top songs were released in March, May and June. 

```{r}
monthly_distrib <- spotify_data_clean %>%
  group_by(release_month) %>%
  summarize(
    count = n()
  ) 

monthly_distrib <- monthly_distrib %>%
  mutate(top3 = rank(-count) %in% 1:3)
  
ggplot(data = monthly_distrib) +
  aes(x = release_month, 
      y = count, 
      fill= top3) +
  geom_bar(stat = 'identity', width=0.8) + 
  labs(
    x = "Month of release",
    y = "Number of songs",
    title = "Most hits were released in March, May and June"
  ) +
  scale_fill_manual(values = c("TRUE" = "#1db954", "FALSE" = "light grey")) +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.position="none")
```
Next we look at the date of the month. We can also notice that most hits were released on the days of 6th, 10th and 13th.

Looking at the two graphs, we can see that 13th of May is the day when most songs were published. Therefore, publishing a song around 13th of May can increase chances of becoming a hit song. However, we must do a more detailed analysis between concluding this.

```{r}
day_distrib <- spotify_data_clean %>%
  group_by(release_day) %>%
  summarize(
    count = n()
  ) 

day_distrib <- day_distrib %>%
  mutate(top3 = rank(-count) %in% 1:3)

ggplot(data = day_distrib) + 
  aes(x = release_day, y = count, fill = top3) +
  geom_bar(stat = 'identity', width=0.8, aes(xmin=1)) + 
  labs(
    x = "Day of release",
    y = "",
    title = "Most hits were released on the 6th, 10th and 13th"
  ) +
  scale_fill_manual(values = c("TRUE" = "#1db954", "FALSE" = "light grey")) +
  coord_cartesian(xlim =c(1, 31)) +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"),
        legend.position="none")

```

How long should a song be?

We want to explore how duration can impact the popularity of a song. Looking at the graph, we can see that popularity increases with duration and most songs are between 2.5 and 3 minutes long. This is not surprising. Very short songs fail to establish a connection between the artist and listener.

```{r}
#Relation between song popularity and duration

ggplot(data = spotify_data_clean) +
  aes(x = duration_min, y = popularity) +
  geom_point() +
  geom_smooth(color="#1db954", se=FALSE) +
  theme_bw() +
  labs(
    x = "Duration (in min)",
    y = "Popularity",
    title = "Most hits are between 2.5 to 3 minutes long"
  )

```

